Deterministic and statistical truncated KdV models for anomalous waves induced by abrupt depth change

### 2018-09-28 LatexNotes/Notes.tex
NM

% Experimental parameters
\newcommand{\lamm}{\lambda^{-}}
\newcommand{\lamp}{\lambda^{+}}
\newcommand{\hm}{h^{-}}
\newcommand{\hp}{h^{+}}
\newcommand{\etastd}{\eta_{\text std}}
% Theory for truncated KdV and stat mech
\newcommand{\uhat}{\hat{u}}
\newcommand{\RR}{\mathbb{R}}
\newcommand{\Real}{\text{Re}}
\newcommand{\Fspace}{\mathscr{F}_{\Lambda}}
\newcommand{\Proj}{P_{\Lambda}}
\newcommand{\sumk}{\sum_{k=1}^{\Lambda}}
\newcommand{\sumn}{\sum_{n=1}^{N}}
\newcommand{\uDir}{u_{\text{Dir}}}
\newcommand{\Gibbs}{\mathcal{G}}

@@

% Table
%^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
\begin{table}%[htbp]
\begin{center}
\caption{Basic experimental parameters and scales.}
\label{paramtable}
\begin{tabular}{l l l}
\hline Parameter & Upstream  value \,\, & Downstream value \\
\hline
peak forcing frequency			& 2 Hz			& N/A \\
depth, $h$						& 12.5 cm			& 3 cm \\
peak wavelength, $\lambda_p$		& 38 cm			& 25 cm \\
peak wavenumber, $k_p$			& 0.17 rad/cm		& 0.25 rad/cm \\
dimensionless depth, $k_p h$\,\,\,		& 2.1			& 0.76 \\
\end{tabular}
\end{center}
\end{table}
 %^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%

@@

The important thing to remember is that Johnson uses these characteristic scales for non-dimensionalizing (characteristic scales denoted by a *)
\begin{align}
\eta^* = \eps h_0 , \quad
x^* = \eps^{-1/2} h_0 , \quad
t^* = \eps^{-1/2} h_0/c
\end{align}
where $h_0$ is a characteristic depth (say the upstream depth), $c = \sqrt{g h_0}$ the corresponding wave speed, and $\eps = a / h_0$ is the ratio of wave amplitude to depth. We want to alter Johnson's non-dimensionalization so that $x$ is scaled on some characteristic wavelength (and $t$ scaled accordingly). Importantly, we need to account for variable depth.

Let's establish some basic scales and notation:
% Table
%^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
\begin{table}[h]%[htbp]
\begin{center}
\caption{Table.}
\label{scale_table}
\begin{tabular}{l l l l}
\hline Description & Notation & Upstream  value \,\, & Downstream value \\
\hline
Characteristic wave amplitude	& $a = \etastd = \mean{\eta^2}^{1/2}$	& varied	& varied \\
Peak forcing frequency		& $f_p$				& 2 Hz		& N/A \\
Depth					& $h^{\pm}$			& 12.5 cm		& 3 cm \\
Peak wavelength			& $\lambda_p^{\pm}$	& 38 cm		& 25 cm \\
%Peak wavenumber			& $k_p$				& 0.17 rad/cm	& 0.25 rad/cm \\
%Dimensionless depth			& $k_p h$				& 2.1		& 0.76
\end{tabular}
\end{center}
\end{table}
 %^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%

@@

There are a few important things to remember when we reformulate this equation:
\begin{enumerate}
\item We must retain Johnson's choice to keep the combination $D^{1/4} \eta$ together because that quantity matches across the abrupt depth change (the raw surface displacement does not match).
\item We want to normalize the standard deviation (or energy) of $\eta$ and therefore want to keep Johnson's choice to scale surface displacement on $a = \eps h_0$.
\item We have to scale the spatial variable (called $x$ or $\xi$) on a characteristic wavelength, which will be a different value on either side of the abrupt depth change. We have to do this so that the assumption of spatial periodicity will be reasonably satisfied on both sides.
\end{enumerate}
To achieve 3, we can use the characteristic wavelength from shallow water theory,
\begin{equation}
\lambda_c = {c}/{f_c} = \sqrt{g d}/f_c
\end{equation}
where $f_c$ is a characteristic forcing frequency and $c = \sqrt{g d}$ is the local wavespeed. It is tempting to simply set $f_c$ to the peak forcing frequency $f_p$, which would result in the characteristic wavelength $\lambda_c$ being the peak wavelength $\lambda_p$. Perhaps a better choice, though, is to let $\lambda_c$ be an integer multiple of $\lambda_p$ to allow scales larger than the peak wavelength (which are certainly present in the experiments). We therefore choose
\begin{equation}
\lambda_c = M \lambda_p = M \sqrt{gd} / f_p
\end{equation}
where $M$ is an integer (on the order of 10 or so). Notice that Johnson's formulation already incorporates the correct scaling of $\sqrt{d}$ to define $\xi$, which will make things convenient. Thus, the main change we need to make is to introduce the dimensionless variable
\begin{equation}
\zeta = \frac{2\pi}{\lambda_c} (\bar{x} - c \bar{t}) 
= \frac{2 \pi \delta}{M \eps^{1/2}} \, \xi
\end{equation}
where
\begin{equation}
\delta = h_0 / \lambda_0 = f_p \sqrt{h_0/g}
\end{equation}
To clean things up a little, lets also introduce a slightly different long-space coordinate that uses the upstream wavelength $\lambda_0$ as a characteristic scale instead of $h_0$,
\begin{equation}
s = \frac{2 \pi \beta}{M \lambda_0} \bar{x} = 2 \pi \beta M^{-1} \delta \eps^{-3/2} X
\end{equation}
where $\beta$ is a small parameter that can be taken to be, for example, $\eps$ or $\delta^2$ or something else.

@@

\subsection{Final formulation}
Equation \eqref{varKdVugly} is essentially in the form we want to use, except that the notation is bad. In particular, we will want to reserve the symbol $H$ for the Hamiltonian. So making superficial changes to the notation gives
\begin{align}
\label{varKdV1}
2 \beta u_s + 3 \eps D^{-7/4} \, u u_x
+ \frac{4 \pi^2 \delta^2}{3 M^2} D^{1/2} \, u_{xxx} = 0
\end{align}
Importantly, $u(x,s)$ is $2\pi$-periodic in $x$ and has normalized energy $\mean{u^2} = 1$.
The definitions of each of the dimensionless variables in terms of the original, physical ones (denoted with a bar) are
\begin{align}
& u = D^{1/4} \, \bar{\eta}/a \\
& x = 2\pi \lambda_c^{-1} (\bar{x} - c \bar{t})  \\
& s = 2 \pi \beta M^{-1} \lambda_0^{-1} \bar{x}
\end{align}
A list of the fixed parameters is
\begin{align}
& \lambda_0 = \sqrt{g h_0}/f_p \quad \text{(upstream wavelength)} \\
& \eps = a / h_0 \quad \text{(dimensionless ratio of wave amplitude to depth)} \\
& \delta = h_0 / \lambda_0 \quad \text{(dimensionless ratio of depth to upstream wavelength)} \\
& \beta	\quad \text{(an unspecified small parameter, usually $\eps$ or $\delta^2$)}
\end{align}
Here is a list of the parameters that vary in space due to the fact that depth $d$ varies in space
\begin{align}
& c = \sqrt{gd} \quad \text{(the wavespeed as it varies in space)} \\
&\lambda_c = M \lambda_p = M \sqrt{gd}/f_p \quad \text{(characteristic wavelength as it varies in space)} \\
&D = d/h_0 \quad \text{(dimensionless depth ratio as it varies in space)} \\	
\end{align}
Note, in the experiments, $\delta = 0.33$ and $\eps$ lies in the range $2.4 \times 10^{-3}$--$2.4 \times 10^{-2}$.

Let's make one more cosmetic change. Divide Eq.~\eqref{varKdV1} by $3 \delta^2$ and set $\beta = 3\delta^2/2$ to get
\begin{align}
\label{varKdV}
u_s + \eps \delta^{-2} D^{-7/4} \, u u_x
+ \mu D^{1/2} \, u_{xxx} = 0
\end{align}
where
\begin{equation}
\mu = \frac{4 \pi^2}{9 M^2}
\end{equation}

Then the Hamiltonian structure is given by
\begin{align}
\label{Hamiltonian}
& H^- = \eps \delta^{-2} H_3 - \mu H_2 \\
& H^+ = \eps \delta^{-2} D_0^{-7/4} H_3 - \mu D_0^{1/2} H_2
\end{align}

Note: $\eps \delta^{-2} = a f_p^2 / g$ which is an interesting parameter. It suggests an equivalence between driving amplitude and driving frequency.

>>>>

% Experimental parameters
\newcommand{\lamm}{\lambda^{-}}
\newcommand{\lamp}{\lambda^{+}}
\newcommand{\hm}{h^{-}}
\newcommand{\hp}{h^{+}}
\newcommand{\etastd}{\eta_{\text std}}
% Theory for truncated KdV and stat mech
\newcommand{\uhat}{\hat{u}}
\newcommand{\RR}{\mathbb{R}}
\newcommand{\Real}{\text{Re}}
\newcommand{\Fspace}{\mathscr{F}_{\Lambda}}
\newcommand{\Proj}{P_{\Lambda}}
\newcommand{\sumk}{\sum_{k=1}^{\Lambda}}
\newcommand{\sumn}{\sum_{n=1}^{N}}
\newcommand{\uDir}{u_{\text{Dir}}}
\newcommand{\Gibbs}{\mathcal{G}}
\newcommand{\ampp}{\mathcal{A}}

@@

% Table
%^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
\begin{table}%[htbp]
\begin{center}
\caption{Basic experimental parameters and scales.}
\label{exptable}
\begin{tabular}{l l l}
\hline Parameter & Upstream  value \,\, & Downstream value \\
\hline
peak forcing frequency, $f_p$\,\,		& 2 Hz			& N/A \\
depth, $h$						& 12.5 cm			& 3 cm \\
peak wavelength, $\lambda_p$		& 38 cm			& 25 cm \\
peak wavenumber, $k_p$			& 0.17 rad/cm		& 0.25 rad/cm \\
dimensionless depth, $k_p h$\,\,\,		& 2.1			& 0.76 \\
\end{tabular}
\end{center}
\end{table}
 %^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%

@@

The important thing to remember is that Johnson uses these characteristic scales for non-dimensionalizing (characteristic scales denoted by a *)
\begin{align}
\eta^* = \eps h_0 , \quad
x^* = \eps^{-1/2} h_0 , \quad
t^* = \eps^{-1/2} h_0/c
\end{align}
where $h_0$ is a characteristic depth (say the upstream depth), $c = \sqrt{g h_0}$ the corresponding wave speed, and $\eps = a / h_0$ is the ratio of wave amplitude to depth. We want to alter Johnson's non-dimensionalization so that $x$ is instead scaled on a characteristic wavelength (and $t$ scaled accordingly). Importantly, we need to account for variable depth.

@@

There are a few important things to remember when we reformulate this equation:
\begin{enumerate}
\item We must retain Johnson's choice to keep the combination $D^{1/4} \eta$ together because that quantity matches across the abrupt depth change (the raw surface displacement does not match).
\item We want to normalize the standard deviation (or energy) of $\eta$ and therefore want to keep Johnson's choice to scale surface displacement on $a = \eps h_0$.
\item We have to scale the spatial variable (called $x$ or $\xi$) on a characteristic wavelength, which will be a different value on either side of the abrupt depth change. We have to do this so that the assumption of spatial periodicity will be reasonably satisfied on both sides.
\end{enumerate}
To achieve 3, we can use the characteristic wavelength from shallow water theory,
\begin{equation}
\lambda_c = {c}/{f_c} = \sqrt{g d}/f_c
\end{equation}
where $f_c$ is a characteristic forcing frequency and $c = \sqrt{g d}$ is the local wavespeed. It is tempting to simply set $f_c$ to the peak forcing frequency $f_p$, which would result in the characteristic wavelength $\lambda_c$ being the peak wavelength $\lambda_p$. Perhaps a better choice, though, is to let $\lambda_c$ be an integer multiple of $\lambda_p$ to allow scales larger than the peak wavelength (which are certainly present in the experiments). We therefore choose
\begin{equation}
\lambda_c = M \lambda_p = M \sqrt{gd} / f_p
\end{equation}
where $M$ is an integer (on the order of 10 or so). Notice that Johnson's formulation already incorporates the correct scaling of $\sqrt{d}$ to define $\xi$, which will make things convenient. Thus, the main change we need to make is to introduce the dimensionless variable
\begin{equation}
\zeta = \frac{2\pi}{\lambda_c} (\bar{x} - c \bar{t}) 
= \frac{2 \pi \delta}{M \eps^{1/2}} \, \xi
\end{equation}
where
\begin{equation}
\delta = h_0 / \lambda_0 = f_p \sqrt{h_0/g}
\end{equation}
To clean things up a little, lets also introduce a slightly different long-space coordinate that uses the upstream wavelength $\lambda_0$ as a characteristic scale instead of $h_0$,
\begin{equation}
s = \frac{2 \pi \beta}{M \lambda_0} \bar{x} = 2 \pi \beta M^{-1} \delta \eps^{-3/2} X
\end{equation}
where $\beta$ is a small parameter that scale like $\eps$ or $\delta^2$ for example.

@@

\subsection{Final formulation of variable-depth KdV}

Equation \eqref{varKdVugly} is essentially in the form we want to use, except that the notation is bad. In particular, we will want to reserve the symbol $H$ for the Hamiltonian. So lets introduce some new variables
\begin{align}
& u = D^{1/4} \, \bar{\eta}/a \\
& x = 2\pi \lambda_c^{-1} (\bar{x} - c \bar{t})  \\
& s = 2 \pi \beta M^{-1} \lambda_0^{-1} \bar{x}
\end{align}
Above, the original physical variables are denoted with a bar. Importantly, $u(x,s)$ is $2\pi$-periodic in $x$ and has normalized energy $\mean{u^2} = 1$.

% Table
%^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
\begin{table}[h]%[htbp]
\begin{center}
\caption{Table of parameters}
\label{paramtable}
\begin{tabular}{l l l l}
\hline \multicolumn{4} { c }{Parameters that are constant throughout the domain} \\
\hline Description & Notation and definition & Units & Value in experiments \\
\hline
Peak forcing frequency		& $f_p$						& frequency	& 2 Hz \\
Characteristic wave amplitude	& $a = \mean{\eta^2}^{1/2}$		& length		& 0.03--0.3 cm \\
Upstream wavelength			& $\lambda_0 = \sqrt{g h_0}/f_p$	& length		& 56 cm \\
Amplitude-to-depth ratio		& $\eps = a / h_0$				& dimensionless	& 0.0024--0.024 \\
Depth-to-wavelength ratio		& $\delta = h_0 / \lambda_0$		& dimensionless		& 0.22 \\
Amplitude parameter			& $\ampp = \eps \delta^{-2} = ag h_0^{-2} f_p^{-2}$	
& dimensionless		& 0.05--0.5\\
Catch-all parameter		& $\mu = 4 \pi^2 / (9 M^2)$			& dimensionless		& 0.27 if $M=4$ \\
\hline \multicolumn{4} { c }{Parameters that change value across ADC} \\
\hline Description & Notation and definition & Value in experiments \\
\hline
Local depth			& $d$					& length \\
Local wavespeed		& $c = \sqrt{gd}$			& speed \\
Local peak wavelength	& $\lambda_p = \sqrt{gd}/f_p$	& length \\
Characteristic wavelength	& $\lambda_c = M \lambda_p$	& length \\
Depth ratio			& $D = d/h_0$				& dimensionless
\end{tabular}
\end{center}
\end{table}
 %^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%

In Table \ref{paramtable} we provide a list of all the relevant parameters needed to convert these dimensional variables to physical ones. Note that the wavelength $\lambda_0$ is predicted from shallow-water theory and so it differs somewhat from the value in Table \ref{exptable} obtained from the dispersion relation. In terms of the new variables,  Eq.~\eqref{varKdVugly} becomes
\begin{align}
\label{varKdV1}
2 \beta u_s + 3 \eps D^{-7/4} \, u u_x
+ \frac{4 \pi^2 \delta^2}{3 M^2} D^{1/2} \, u_{xxx} = 0
\end{align}

Let's clean it up now a little bit. Let us set $\beta = 3 \delta^2/2$ and introduce the parameters
\begin{align}
\ampp = \eps \delta^{-2} = \frac{ag}{h_0^2 f_p^2} , \qquad
\mu = \frac{4 \pi^2}{9 M^2}
\end{align}
The Eq.~\eqref{varKdV1} becomes
\begin{align}
\label{varKdV}
u_s + \ampp D^{-7/4} \, u u_x
+ \mu D^{1/2} \, u_{xxx} = 0
\end{align}

Then the Hamiltonian structure is given by
\begin{align}
\label{Hamiltonian}
& H^- = \ampp H_3 - \mu H_2 \\
& H^+ = \ampp D_0^{-7/4} H_3 - \mu D_0^{1/2} H_2
\end{align}
